---
title: "Data stories: Climate change"
date: 2024-11-01
categories:
    - climate change
    - data science
---

It's impossible to ignore the big decisions that Americans will be making next Tuesday.
But neither campaign is talking about climate change in a serious way.
I have my own thoughts about why, but let's look at some data rather than pontificate.

I found that there is an Emissions Database for Global Atmospheric Research (EDGAR) site that collects data on these questions. 
Here is the site for emissions data from 1970 to 2024 for a large set of countries:

<https://data.jrc.ec.europa.eu/dataset/74f34d14-5819-4954-8b61-b11ce3d4ca48>

The following code downloads the data.

```{r}
url <- "https://jeodpp.jrc.ec.europa.eu/ftp/jrc-opendata/EDGAR/datasets/EDGAR_2024_report/EDGAR_2024_GHG_booklet_2024_fossilCO2only.xlsx"

download.file(url, destfile = "data/edgar-co2-2024.xlsx", mode = "wb")
```

This is an Excel spreadsheet, so the [readxl](https://readxl.tidyverse.org) package will be helpful.

Some poking around in the Excel spreadsheet indicates two tabs that are especially relevant `fossil_CO2_per_GDP_by_country` and `fossil_CO2_per_capita_by_country`.

```{r}
sheets_avail <- readxl::excel_sheets("data/edgar-co2-2024.xlsx")
sheets_avail
```

```{r}
CO2_per_capita <- readxl::read_excel("data/edgar-co2-2024.xlsx", sheet = sheets_avail[7])
str(CO2_per_capita)
```

We need to make this longer if we want to use Tidyverse-style plotting commands.

```{r}
CO2_per_capita_long <- CO2_per_capita |>
  tidyr::pivot_longer(
    cols = (starts_with("19") |
              starts_with("20")),
    names_to = "year",
    values_to = "emissionsPerCapita"
  ) |>
  # Make the year a properly formatted YYYY-MM-DD string for parsing
  dplyr::mutate(year = lubridate::as_date(paste0(year, "-12-31")))
```

Now we can make our first attempt at a plot.

```{r}
#| label: fig-co2-emissions-1970-2023-big-economies
#| fig-cap: "CO2 emissions per capita 1970-2023 from the Emissions Database for Global Atmospheric Research (EDGAR) database."
library(ggplot2)
CO2_per_capita_long |>
  dplyr::filter(
    Country %in% c(
      "United States",
      "China",
      "Japan",
      "Germany",
      "France and Monaco",
      "Canada",
      "United Kingdom",
      "Taiwan",
      "South Korea",
      "Australia"
    )
  ) |>
  ggplot2::ggplot() +
  aes(year, emissionsPerCapita, color = Country, group = Country) +
  geom_point() +
  geom_line() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90))
```

Let's look at the top countries in 2023.

```{r}
top_2023 <- CO2_per_capita_long |>
  dplyr::filter(year == "2023-12-31") |>
  dplyr::arrange(desc(emissionsPerCapita))

top_20_2023 <- top_2023[1:20,]
top_20_countries_2023 <- top_20_2023$Country

top_20_2023 |>
  dplyr::select(Country, emissionsPerCapita) |>
  kableExtra::kable(format = 'html') |>
  kableExtra::kable_classic()
```

Let's see how these countries fared since 1970.

```{r}
CO2_per_capita_long |>
  dplyr::filter(Country %in% top_20_2023$Country) |>
  ggplot2::ggplot() +
  aes(year, emissionsPerCapita, color = Country, group = Country) +
  geom_point() +
  geom_line() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90))
```

Now, let's see how the top 20 in 1970 have changed.

```{r}
top_1970 <- CO2_per_capita_long |>
  dplyr::filter(year == "1970-12-31") |>
  dplyr::arrange(desc(emissionsPerCapita))

top_20_1970 <- top_1970[1:20,]
top_20_countries_1970 <- top_20_1970$Country

top_20_1970 |>
  dplyr::select(Country, emissionsPerCapita) |>
  kableExtra::kable(format = 'html') |>
  kableExtra::kable_classic()
```

```{r}
CO2_per_capita_long |>
  dplyr::filter(Country %in% top_20_1970$Country) |>
  ggplot2::ggplot() +
  aes(year, emissionsPerCapita, color = Country, group = Country) +
  geom_point() +
  geom_line() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90))
```

So, the top emitters per capita in 1970 have made substantial progress in reducing emissions.
The top emitters in 2023 also made substantial reductions from 1970 to 2023.

The East Asian countries that have have increased emissions per capita since 1970 (China, Taiwan, Japan, South Korea) have joined the mid-ranks of the other wealthy countries in emissions per capita.

This all strikes me as very interesting, and largely positive news.
The data story I would tell is that the world is getting richer, and that per capita CO2 emissions are falling into a similar range, even among countries that have grown quickly over the past 50 years.
We just need to do more, especially those of us living in wealthy, large countries who can.
